---
title: "HW 3: Visualization and EDA"
author: "Emma Gorin"
date: "October 2020"
output: github_document
---

```{r setup, message = FALSE}
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1
This problem uses the Instacart data. DO NOT include this dataset in your local data directory; instead, load the data from the p8105.datasets using:

```{r load_instacart}
library(p8105.datasets)
data("instacart")
```

The goal is to do some exploration of this dataset. To that end, write a short description of the dataset, noting the size and structure of the data, describing some key variables, and giving illustrative examples of observations. 

**The instacart dataset includes data about grocery orders placed through Instacart. The dataset includes `r ncol(instacart)` columns and `r nrow(instacart)` observations, each representing a single item included in an order. Variables include both details about the order (such as order id number, days since prior order, hour of the day at which the order was placed) and details about the product (such as product name, the aisle and department in which the product can be found, and whether the product was re-ordered.) For instance, we can see that the order with order id #`r instacart[68,1]` included `r instacart[68, 11]` which can be found in the `r instacart[68, 14]` aisle and which was reordered.**

Then, do or answer the following (commenting on the results of each):

* How many aisles are there, and which aisles are the most items ordered from?

```{r}
most_pop_aisles.df = instacart %>%
  group_by(aisle) %>% 
  summarize(n_obs = n()) %>% 
  arrange(desc(n_obs))
```

**There are 134 aisles. Most items are ordered from the `r most_pop_aisles.df[1,1]` aisle with `r most_pop_aisles.df[1,2]` items ordered, closely followed by the `r most_pop_aisles.df[2,1]` aisle with `r most_pop_aisles.df[2,2]` items ordered. Distant 3rd, 4th and 5th places are held by `r most_pop_aisles.df[3,1]`, `r most_pop_aisles.df[4,1]`, and `r most_pop_aisles.df[5,1]`.**

* Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

```{r}
instacart %>% 
  ungroup() %>% 
  mutate(aisle = factor(aisle)) %>% 
  group_by(aisle) %>% 
  mutate(count_aisle = n()) %>% 
  filter(count_aisle > 10000) %>% 
  select(aisle, count_aisle, everything()) %>% 
  ggplot(aes(x = fct_rev(fct_infreq(aisle)))) +
  geom_bar() +
  coord_flip() +
  labs(
    title = "Most popular aisles in Instacart dataset",
    y = "Count of items ordered",
    x = "Aisle name"
  )
```

* Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r}
instacart %>% 
  ungroup() %>% 
  filter(aisle == "baking ingredients" | aisle == "dog food care" | aisle == "packaged vegetables fruits") %>% 
  group_by(aisle, product_name) %>% 
  summarize(times_ordered = n()) %>% 
  filter(min_rank(desc(times_ordered)) < 4) %>% 
  select(aisle, product_name, times_ordered) %>% 
  arrange(aisle, desc(times_ordered))
```

* Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r}
instacart %>% 
  filter(product_name == "Pink Lady Apples" | product_name == "Coffee Ice Cream") %>%
  group_by(product_name, order_dow) %>% 
  summarize(time_of_order = mean(order_hour_of_day)) %>% 
  mutate(day_of_the_week = recode(order_dow,
                                  "0"="Sunday",
                                  "1"="Monday",
                                  "2"="Tuesday",
                                  "3"="Wednesday",
                                  "4"="Thursday",
                                  "5"="Friday",
                                  "6"="Saturday")) %>% 
  pivot_wider(id_cols = product_name, names_from = day_of_the_week, values_from = time_of_order) 
```

**Looks like pink lady apples are typically ordered a little earlier in the day than coffee ice cream. Not too surprising.**

## Problem 2
Accelerometers have become an appealing alternative to self-report techniques for studying physical activity in observational studies and clinical trials, largely because of their relative objectivity. During observation periods, the devices measure “activity counts” in a short period; one-minute intervals are common. Because accelerometers can be worn comfortably and unobtrusively, they produce around-the-clock observations.

This problem uses five weeks of accelerometer data collected on a 63 year-old male with BMI 25, who was admitted to the Advanced Cardiac Care Center of Columbia University Medical Center and diagnosed with congestive heart failure (CHF). The data can be downloaded here. In this spreadsheet, variables activity.* are the activity counts for each minute of a 24-hour day starting at midnight.

* Load, tidy, and otherwise wrangle the data. Your final dataset should include all originally observed variables and values; have useful variable names; include a weekday vs weekend variable; and encode data with reasonable variable classes. Describe the resulting dataset (e.g. what variables exist, how many observations, etc).

```{r}
accelerom = read_csv("./data/accel_data.csv") %>% 
  janitor::clean_names() %>% 
  mutate(weekday_or_end = recode(day, "Sunday" = "weekend",
                                  "Monday" = "weekday",
                                  "Tuesday" = "weekday",
                                  "Wednesday" = "weekday",
                                  "Thursday" = "weekday",
                                  "Friday" = "weekday",
                                  "Saturday" = "weekend")) %>% 
  mutate(day = factor(day)) %>% 
  mutate(weekday_or_end = factor(weekday_or_end)) %>% 
  select(week, day_id, day, weekday_or_end, everything()) %>% 
  pivot_longer(cols = starts_with("activity"), names_to = "activity_minute", values_to = "activity_count")
```

**The accelerometer dataset includes `r nrow(accelerom)` observations, one per minute over the course of a five week period. Variables include week, day, whether it is a weekday or weekend, and "activity count" as recorded by an accelerometer.**

* Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate accross minutes to create a total activity variable for each day, and create a table showing these totals. Are any trends apparent?

```{r}

```

* Accelerometer data allows the inspection activity over the course of the day. Make a single-panel plot that shows the 24-hour activity time courses for each day and use color to indicate day of the week. Describe in words any patterns or conclusions you can make based on this graph.

```{r}

```


## Problem 3

This problem uses the NY NOAA data. DO NOT include this dataset in your local data directory; instead, load the data from the p8105.datasets package using:

```{r}
library(p8105.datasets)
data("ny_noaa")
```

The goal is to do some exploration of this dataset. To that end, write a short description of the dataset, noting the size and structure of the data, describing some key variables, and indicating the extent to which missing data is an issue. Then, do or answer the following (commenting on the results of each):

* Do some data cleaning. Create separate variables for year, month, and day. Ensure observations for temperature, precipitation, and snowfall are given in reasonable units. For snowfall, what are the most commonly observed values? Why?

```{r}

```

* Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers?

```{r}

```

* Make a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option); and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.

```{r}

```





